Data Pre-Processing/Cleaning Script

### 2023-03-14 code/analysis_code/exploratory_analysis.qmd
update processing and exploratory analysis files

## Analysis Data
```{r}
# Path to Data
data_location <- here(
  "data","processed_data","processed_analysis_data.rds")
# Import Data
analysis_data_final <- readRDS(data_location)
```

@@

## Closer Look at Bacterial Etiologies for Outbreaks
ggplot(training_data_final %>% 
         select(Pathogen_Type, Simplified_Etiology, Hospitalizations) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + 
  labs(x = "Bacteria", y = "Hospitalizations", title = "Bar Chart of Bacteria Frequency")

@@

# Food Type and Hospitalizations
ggplot(training_data_final %>% 
         filter(`IFSAC Category` != "Multiple") %>% 
         filter(`IFSAC Category` != "Other"), 
       aes(x = `IFSAC Category`, y = Hospitalizations)) + 
  geom_col() + 
  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")

@@

# Locations and Hospitalizations
Location_v_Hospitalization <- ggplot(training_data_final, aes(x = Location, y = Hospitalizations)) +
  geom_col() + 
  labs(x = "Locations", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")+
  scale_x_discrete(guide=guide_axis(n.dodge=3))

@@

## Analysis Data
```{r}
# Illnesses Caused by Pathogen Type Over Time
ggplot(analysis_data_final %>% 
         group_by(Year, Pathogen_Type, Illnesses) %>% 
         count() %>% 
         mutate(true_illness_count = n * Illnesses), 
       aes(x = Year, y = true_illness_count, group = Pathogen_Type, color = Pathogen_Type)) + 
  geom_point() + 
  geom_jitter() + 
  labs(x = "Year", y = "Cases", title = "Scatter Plot of Illness by Year and Pathogen Type")

@@

# Pathogen Type + Etiology vs Illnesses Caused
ggplot(analysis_data_final %>% 
         group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% 
         count() %>% 
         mutate(true_illness_count = n * Illnesses), 
       aes(x = Pathogen_Type, y = Illnesses, group = Simplified_Etiology, fill = Simplified_Etiology)) + 
  geom_col() + 
  labs(x = "Pathogen Type", y = "Cases", title = "Stacked Bar Chart of Illnesses by Pathogen Type and Etiology")

@@

## Closer Look at Bacterial Etiologies
ggplot(analysis_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency")

@@

state_v_illness_analysis <- analysis_data_final %>% group_by(State) %>% mutate(all_illnesses = sum(Illnesses)) %>% count(all_illnesses) %>% select(State, all_illnesses) %>% arrange(desc(all_illnesses))

saveRDS(state_v_illness_analysis, file = table_path)

@@

# Food Type and Illnesses
ggplot(analysis_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + geom_col() + labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")+
  scale_x_discrete(guide = guide_axis(n.dodge=3))

@@

# Locations and Illnesses
ggplot(analysis_data_final, aes(x = Location, y = Illnesses)) +
  geom_col() + 
  labs(x = "Locations", y = "Illnesses", title = "Illnesses by Locations")+
  scale_x_discrete(guide=guide_axis(n.dodge=3))

>>>>

## Analysis Data
```{r}
# Path to Data
data_location <- here(
  "data","processed_data","processed_testing_data.rds")
# Import Data
analysis_data_final <- readRDS(data_location)
```

@@

## Closer Look at Bacterial Etiologies for Outbreaks
ggplot(training_data_final %>% 
         select(Pathogen_Type, Simplified_Etiology, Hospitalizations, `IFSAC Category`) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology,y=`Hospitalizations`))+ 
  geom_col() + 
  labs(x = "Bacteria", y = "Hospitalizations", title = "Bar Chart of Bacteria Frequency")+
  coord_flip()

@@

# Food Type and Hospitalizations
ggplot(training_data_final %>% 
         filter(`IFSAC Category` != "Multiple") %>% 
         filter(`IFSAC Category` != "Other"), 
       aes(x = `IFSAC Category`, y = Hospitalizations)) + 
  geom_col() + 
  labs(x = "Food Type", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")+
  coord_flip()

@@

# Locations and Hospitalizations
Location_v_Hospitalization <- ggplot(training_data_final, aes(x = Location, y = Hospitalizations)) +
  geom_col() + 
  labs(x = "Locations", y = "Hospitalizations", title = "Hospitalizations by Food Type, Excluding 'Multiple' and 'Other'")+
  coord_flip()

@@

## Testing Data
```{r}
# Illnesses Caused by Pathogen Type Over Time
ggplot(testing_data_final %>% 
         group_by(Year, Pathogen_Type, Illnesses) %>% 
         count() %>% 
         mutate(true_illness_count = n * Illnesses), 
       aes(x = Year, y = true_illness_count, group = Pathogen_Type, color = Pathogen_Type)) + 
  geom_point() + 
  geom_jitter() + 
  labs(x = "Year", y = "Cases", title = "Scatter Plot of Illness by Year and Pathogen Type")

@@

# Pathogen Type + Etiology vs Illnesses Caused
ggplot(testing_data_final %>% 
         group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% 
         count() %>% 
         mutate(true_illness_count = n * Illnesses), 
       aes(x = Pathogen_Type, y = Illnesses, group = Simplified_Etiology, fill = Simplified_Etiology)) + 
  geom_col() + 
  labs(x = "Pathogen Type", y = "Cases", title = "Stacked Bar Chart of Illnesses by Pathogen Type and Etiology")

@@

## Closer Look at Bacterial Etiologies
ggplot(testing_data_final %>% group_by(Pathogen_Type, Simplified_Etiology, Illnesses) %>% filter(Pathogen_Type == "Bacteria"), aes(x = Simplified_Etiology)) + geom_bar() + labs(x = "Bacteria", y = "Cases", title = "Bar Chart of Bacteria Frequency")+
  coord_flip()

@@

state_v_illness_testing <- testing_data_final %>% group_by(State) %>% mutate(all_illnesses = sum(Illnesses)) %>% count(all_illnesses) %>% select(State, all_illnesses) %>% arrange(desc(all_illnesses))

saveRDS(state_v_illness_testing, file = table_path)

@@

# Food Type and Illnesses
ggplot(testing_data_final %>% filter(`IFSAC Category` != "Multiple") %>% filter(`IFSAC Category` != "Other"), aes(x = `IFSAC Category`, y = Illnesses)) + geom_col() + labs(x = "Food Type", y = "Illnesses", title = "Illnesses by Food Type, Excluding 'Multiple' and 'Other'")+
  coord_flip()

@@

# Locations and Illnesses
ggplot(testing_data_final, aes(x = Location, y = Illnesses)) +
  geom_col() + 
  labs(x = "Locations", y = "Illnesses", title = "Illnesses by Locations")+
  coord_flip()

### 2023-03-14 code/processing_code/processingfile_v1.qmd
update processing and exploratory analysis files

data_location <- here(
  "data","raw_data","NationalOutbreakPublicDataTool.xlsx")
analysis_data <- readxl::read_excel(data_location)
```

@@

## Analysis Data
```{r}
str(analysis_data)
summary(analysis_data)

@@

## Training Data
```{r}
training_data_fin <- training_data %>%
  select(!c(Ingredient, `Serotype/Genotype`)) %>%
  filter(grepl("Confirmed", Status))
```
# Data Manipulation
## Training Data
### Cleaning the "Species" Column
```{r}
unique(training_data_fin$Species)
# Separate the rows for coninfections
training_data_fin=training_data_fin%>%
  separate_rows(Species,sep=";")
# Create lists for pathogen types
Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia","Streptococcus")
Pathogen_Parasite <- c("Cryptosporidium", "Cyclospora", "Entamoeba", "Giardia", "Parasite", "Toxoplasma", "Trichinella")
Pathogen_Virus <- c("Astrovirus", "Norovirus", "Rotavirus","Hepatitis","Virus")
Pathogen_Poison <- c("Poison", "Toxin", "Metal","Pesticide","Ciguatoxin","Scombroidtoxin")
# simplify "Escherichia coli, Shiga toxin-producing" to "Escherichia coli"
# Create two new columns: Pathogen_type & Simplified_Etiology
training_data_fin=training_data_fin%>%
  mutate(Pathogen_Type = case_when(
    grepl(paste(Pathogen_Virus, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Virus",
    grepl(paste(Pathogen_Bacteria, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Bacteria",
    grepl(paste(Pathogen_Parasite, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Parasite",
    grepl(paste(Pathogen_Poison, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Poison"
    )) %>%
  mutate(Simplified_Etiology = case_when(
    Pathogen_Type == "Bacteria" ~ str_extract(Species, 
      "Bacillus|Brucella|Campylobacter|Clostridium|Staphylococcus|Escherichia|Listeria|Salmonella|Vibrio|Cholera|Enterococcus|Bacterium|Shigella|Yersinia|Streptococcus"),
    Pathogen_Type == "Parasite" ~ str_extract(Species, "Cryptosporidium|Cyclospora|Entamoeba|Giardia|Parasite|Toxoplasma|Trichinella"),
    Pathogen_Type == "Virus" ~ str_extract(Species, "Astrovirus|Norovirus|Rotavirus|Hepatitis"),
    Pathogen_Type == "Poison" ~ str_extract(Species, "Poison|Toxin|Heavy|pesticide|Ciguatoxin|Scombroid")
    )) %>%
      mutate(Simplified_Etiology = case_when(
        Simplified_Etiology == "Escherichia" ~ "E. coli",
        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
        TRUE ~ Simplified_Etiology))
```

### Column for Location
```{r}
# Checking Location Names and Separate Multiple Names
unique(training_data_fin$Location)
training_data_fin=training_data_fin %>%
 separate_rows(Location,sep="; ")

@@

analysis_data_fin <- analysis_data_fin %>%
  mutate(Pathogen_Type = case_when(
    grepl(paste(Pathogen_Poison, 
                collapse = "|"), Etiology) ~ "Poison",
    grepl(paste(Pathogen_Bacteria, 
                collapse = "|"), Etiology) ~ "Bacteria",
    grepl(paste(Pathogen_Parasite, 
                collapse = "|"), Etiology) ~ "Parasite",
    grepl(paste(Pathogen_Virus, 
                collapse = "|"), Etiology) ~ "Virus"
    )) %>%
  mutate(Simplified_Etiology = case_when(
    Pathogen_Type == "Bacteria" ~ str_extract(Etiology, 
      "Bacillus|Brucella|Campylobacter|Clostridium|Staphylococcus|Escherichia|Listeria|Salmonella|Vibrio|Cholera|Enterococcus|Bacterium|Shigella|Yersinia"),
    Pathogen_Type == "Parasite" ~ str_extract(Etiology, "Cryptosporidium|Cyclospora|Entamoeba|Giardia|Parasite|Toxoplasma|Trichinella"),
    Pathogen_Type == "Virus" ~ str_extract(Etiology, "Astrovirus|Norovirus|Rotavirus"),
    Pathogen_Type == "Poison" ~ str_extract(Etiology, "Poison|Toxin|Heavy")
    )) %>%
      mutate(Simplified_Etiology = case_when(
        Simplified_Etiology == "Escherichia" ~ "E. coli",
        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
        TRUE ~ Simplified_Etiology))
```
###Column for Location
```{r}
analysis_data_fin_re <- analysis_data_fin %>%
  rename(Location=Setting)%>%
  mutate(Location=case_when(
    grepl("Restaurant",Location)~"Restaurant",TRUE~Location))%>%
    separate_rows(Location,sep="; ")
analysis_data_fin_re
```

>>>>

data_location <- here(
  "data","raw_data","NationalOutbreakPublicDataTool.xlsx")
testing_data <- readxl::read_excel(data_location)
```

@@

## Testing Data
```{r}
str(analysis_data)
summary(analysis_data)

@@

## Training Data
```{r}
training_data_fin <- training_data %>%
  select(!c(Ingredient, `Serotype/Genotype`)) %>%
  filter(grepl("Confirmed", Status))
```
# Data Manipulation
## Training Data
### Cleaning the "Species" Column
```{r}
unique(training_data_fin$Species)
# Separate the rows for coninfections
training_data_fin=training_data_fin%>%
  separate_rows(Species,sep=";")
# Create lists for pathogen types
Pathogen_Bacteria <- c("Bacillus", "Brucella", "Campylobacter", "Clostridium", "Staphylococcus", "Escherichia", "Listeria", "Salmonella", "Vibrio", "Cholera", "Enterococcus", "Bacterium", "Shigella", "Yersinia","Streptococcus")
Pathogen_Parasite <- c("Cryptosporidium", "Cyclospora", "Entamoeba", "Giardia", "Parasite", "Toxoplasma", "Trichinella")
Pathogen_Virus <- c("Astrovirus", "Norovirus", "Rotavirus","Hepatitis","Virus")
Pathogen_Poison <- c("Poison", "Toxin", "Metal","Pesticide","Ciguatoxin","Scombroidtoxin")
# simplify "Escherichia coli, Shiga toxin-producing" to "Escherichia coli"
# Create two new columns: Pathogen_type & Simplified_Etiology
training_data_fin=training_data_fin%>%
  mutate(Pathogen_Type = case_when(
    grepl(paste(Pathogen_Virus,
                collapse="|"), Species,ignore.case = TRUE) ~ "Virus",
    grepl(paste(Pathogen_Bacteria, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Bacteria",
    grepl(paste(Pathogen_Parasite, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Parasite",
    grepl(paste(Pathogen_Poison, 
                collapse = "|"), Species,ignore.case = TRUE) ~ "Poison"
    )) %>%
  mutate(Simplified_Etiology = word(Species,1))%>%
      mutate(Simplified_Etiology = case_when(
        Simplified_Etiology == "Escherichia" ~ "E. coli",
        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
        TRUE ~ Simplified_Etiology))
```

### Column for Location
```{r}
# Checking Location Names and Separate Multiple Names
unique(training_data_fin$Location)
training_data_fin=training_data_fin %>%
 separate_rows(Location,sep="; ")
```

@@

analysis_data_fin <- analysis_data_fin %>%
  mutate(Pathogen_Type = case_when(
    grepl(paste(Pathogen_Poison, 
                collapse = "|"), Etiology) ~ "Poison",
    grepl(paste(Pathogen_Bacteria, 
                collapse = "|"), Etiology) ~ "Bacteria",
    grepl(paste(Pathogen_Parasite, 
                collapse = "|"), Etiology) ~ "Parasite",
    grepl(paste(Pathogen_Virus, 
                collapse = "|"), Etiology) ~ "Virus"
    )) %>%
  mutate(Simplified_Etiology = word(Etiology,1)) %>%
      mutate(Simplified_Etiology = case_when(
        Simplified_Etiology == "Escherichia" ~ "E. coli",
        Simplified_Etiology == "Heavy" ~ "Heavy Metals",
        TRUE ~ Simplified_Etiology))
```
###Column for Location
```{r}
analysis_data_fin_re <- analysis_data_fin %>%
  rename(Location=Setting)%>%
  mutate(Location=case_when(
    grepl("Restaurant",Location)~"Restaurant",TRUE~Location))%>%
    separate_rows(Location,sep="; ")
analysis_data_fin_re
```
